---
title: "Tables Chapter 4 v2"
author: "Fabian Welti"
date: "2024-03-30"
output: html_document
---

Setup:
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
library(kableExtra)
library(xts)
library(ggplot2)
library(zoo)
library(dplyr)
library(TTR)
library(broom)
library(moments)
library(tibble)
library(tidyr)
library(dplyr)
library(readxl)
options(scipen = 999)


```

Define the font 
```{css, echo=FALSE}
.times-new-roman {
  font-family: "Times New Roman", Times, serif !important;
}
```


Read in all the necessary data: 
```{r}

#WD ####
setwd("/Users/lucaswelti/Documents/Data BA Thesis /Futures/2006-2013")

#Packages ####
library(readr)
library(dplyr)

#Combine the individual files for the monthly VIX futures ####

# List all CSV files in WD
csv_files_06_13 <- list.files(pattern = "^CFE_[FGHJKMNQUVXZ]\\d{2}_VX\\.csv$")
print(csv_files_06_13)

# Read and combine the files into one data frame
all_data_06_13 <- lapply(csv_files_06_13, read_csv)

# Combine all data frames into one
combined_data_06_13 <- bind_rows(all_data_06_13)
combined_data_06_13 <- combined_data_06_13 %>%
  rename(Date = `Trade Date`) #rename column

# Rescale the Values before 26 March 2007
  # Convert the 'Date' column to a Date type 
combined_data_06_13$Date <- as.Date(combined_data_06_13$Date,format = "%m/%d/%Y")
  # Set the rescale date
rescale_date <- as.Date("2007-03-26")
  # Rescale the data if the 'Date' is before the 'rescale_date'
combined_data_06_13 <- combined_data_06_13 %>%
  mutate(across(c(Open, High, Low, Close, Settle, Change), 
                ~ if_else(Date < rescale_date, . / 10, .)))


# Check the structure and summary of the combined data frame
str(combined_data_06_13)
summary(combined_data_06_13)

# Write the combined data frame to a CSV file
write_csv(combined_data_06_13, "combined_vix_futures_06_13.csv")


#WD ####
setwd("/Users/lucaswelti/Documents/Data BA Thesis /Futures/2014-2024/Monthly ")

#Packages ####
library(readr)
library(dplyr)
library(stringr)


#Combine the individual files for the monthly VIX futures ####

# Define a vector with the year suffixes according to the provided mapping
year_suffixes_14_24 <- c("14" = "2014", "5" = "2015", "6" = "2016", "7" = "2017", 
                   "8" = "2018", "9" = "2019", "0" = "2020", "1" = "2021", 
                   "2" = "2022", "3" = "2023", "4" = "2024")

# Generate filenames for all possible combinations
file_patterns_14_24 <- paste0("CFE_VX_[FGHJKMNQUVXZ](", paste(names(year_suffixes_14_24), collapse = "|"), ")(-\\d+)?\\.csv")

# List all CSV files that match the generated patterns
csv_files_14_24 <- list.files(pattern = file_patterns_14_24)
print(csv_files_14_24)

# Read and combine the files into one data frame
all_data_14_24 <- lapply(csv_files_14_24, function(file_name) {
  # Extract the year code from the file name
  year_code <- str_extract(file_name, paste(names(year_suffixes_14_24), collapse = "|"))
  
  # Read the CSV file
  data_14_24 <- read_csv(file_name)
  
  # Add a Year column based on the extracted year code
  data_14_24$Year <- year_suffixes_14_24[year_code]
  
  return(data_14_24)
})

# Combine all
combined_data_14_24 <- bind_rows(all_data_14_24)
combined_data_14_24 <- combined_data_14_24 %>%
  rename(Date = `Trade Date`)  # Rename 'Trade Date' to 'Date'
combined_data_14_24$Date <- as.Date(combined_data_14_24$Date, format = "%Y-%m-%d") 


# Check the structure and summary of the combined data frame
str(combined_data_14_24)
summary(combined_data_14_24)

# Write the combined data frame to a CSV file
write_csv(combined_data_06_13, "combined_vix_futures_14_24.csv")


library(readr)
library(dplyr)
library(stringr)
library(lubridate)
library(timeDate)

#Merge all the data for VIX futures from 2006-2024 ####
vix_futures_06_24 <- bind_rows(combined_data_06_13, combined_data_14_24)

vix_futures_06_24 <- vix_futures_06_24 %>%
  select(-ncol(vix_futures_06_24)) #remove the last column 

# arrange by ascending order of date
vix_futures_06_24 <- vix_futures_06_24 %>%
  arrange(Date)

# safe as a CSV file 
write.csv(vix_futures_06_24, "vix futures from 06 until 24", row.names = FALSE)




#Manipulate the Data ####


# new column with futures expiration month
vix_futures_06_24 <- vix_futures_06_24 %>%
  mutate(`Futures expiration month` = str_extract(Futures, "\\(([^)]+)\\)")) # The regular expression "\\(([^)]+)\\)" matches any character between parentheses

# new column with the futures expiration year 
vix_futures_06_24 <- vix_futures_06_24 %>%
  mutate(`Expiration Year` = str_extract(`Futures expiration month`, "\\d{2}(?=\\))"))

# in Futures expiration month only keep the month
vix_futures_06_24 <- vix_futures_06_24 %>%
  mutate(`Futures expiration month` = str_extract(`Futures expiration month`, "[A-Za-z]+"))

# Change the abbreviation of the month name to the full name 
vix_futures_06_24 <- vix_futures_06_24 %>%
  mutate(`Futures expiration month` = case_when(
    `Futures expiration month` == "Jan" ~ "01",
    `Futures expiration month` == "Feb" ~ "02",
    `Futures expiration month` == "Mar" ~ "03",
    `Futures expiration month` == "Apr" ~ "04",
    `Futures expiration month` == "May" ~ "05",
    `Futures expiration month` == "Jun" ~ "06",
    `Futures expiration month` == "Jul" ~ "07",
    `Futures expiration month` == "Aug" ~ "08",
    `Futures expiration month` == "Sep" ~ "09",
    `Futures expiration month` == "Oct" ~ "10",
    `Futures expiration month` == "Nov" ~ "11",
    `Futures expiration month` == "Dec" ~ "12",
    TRUE ~ as.character(`Futures expiration month`)  # Default case to keep the original value
  ))

# Change the abbreviation of the year name to the full name 
vix_futures_06_24 <- vix_futures_06_24 %>%
  mutate(`Expiration Year` = case_when(
    `Expiration Year` == "06" ~ "2006",
    `Expiration Year` == "07" ~ "2007",
    `Expiration Year` == "08" ~ "2008",
    `Expiration Year` == "09" ~ "2009",
    `Expiration Year` == "10" ~ "2010",
    `Expiration Year` == "11" ~ "2011",
    `Expiration Year` == "12" ~ "2012",
    `Expiration Year` == "13" ~ "2013",
    `Expiration Year` == "14" ~ "2014",
    `Expiration Year` == "15" ~ "2015",
    `Expiration Year` == "16" ~ "2016",
    `Expiration Year` == "17" ~ "2017",
    `Expiration Year` == "18" ~ "2018",
    `Expiration Year` == "19" ~ "2019",
    `Expiration Year` == "20" ~ "2020",
    `Expiration Year` == "21" ~ "2021",
    `Expiration Year` == "22" ~ "2022",
    `Expiration Year` == "23" ~ "2023",
    `Expiration Year` == "24" ~ "2024",
    TRUE ~ paste0("20", `Expiration Year`)  # For years beyond the range specified
  ))







vix_futures_06_24 <- vix_futures_06_24 %>%
  mutate(`Expiration month and year` = paste(`Futures expiration month`, `Expiration Year`, sep = "-"))






#Calculate the expiration month ####
calculate_expiration_date <- function(date_string) {
  # Parse the month and year from the date string
  parts <- strsplit(date_string, "-")[[1]]
  month <- as.numeric(parts[1])
  year <- as.numeric(parts[2])
  
  # Move to the first day of the next month
  if(month == 12) {
    month <- 1
    year <- year + 1
  } else {
    month <- month + 1
  }
  
  date <- as.Date(paste(year, month, "01", sep = "-"))
  
  # Calculate the third Friday of the next month
  # Add days until it's the first Friday
  while (format(date, "%A") != "Friday") {
    date <- date + 1
  }
  # Then add 14 days to get to the third Friday
  third_friday <- date + 14
  
  # Subtract 30 days to find the expiration date which should be a Wednesday
  expiration_date <- third_friday - 30
  
  # If it's not a business day (Saturday or Sunday), adjust to the previous business day
  while (format(expiration_date, "%u") %in% c("6", "7")) {
    expiration_date <- expiration_date - 1
  }
  
  return(expiration_date)
}

# Apply the function to each row
vix_futures_06_24$ExpirationDate <- as.Date(sapply(vix_futures_06_24$`Expiration month and year`, calculate_expiration_date), origin = "1970-01-01")


# Account manually for some exeptions because of holidays
vix_futures_06_24 <- vix_futures_06_24 %>%
  mutate(ExpirationDate = case_when(
    ExpirationDate == as.Date('2007-11-21') ~ as.Date('2007-11-20'),
    ExpirationDate == as.Date('2024-06-19') ~ as.Date('2024-06-18'),
    TRUE ~ ExpirationDate
  ))



setwd("~/Documents/Data BA Thesis ")
#Step 1: load and clean historical S&P 500 and VIX data ####
vix_data <- read_excel("VIX_History 1990-2024.xlsx", skip = 1)

colnames(vix_data) <- c("Date", "Open", "High", "Low", "Close", "Returns")

vix_data$Date <- as.Date(vix_data$Date, format = "%m/%d/%Y") #converting the Date column 



#Step 2: Calculate the daily Returns for SPX and VIX ####

vix_data$Returns <- log(vix_data$Close/lag(vix_data$Close))
vix_data <- na.omit(vix_data) # Remove NA values 

```


Code for summary stats for levels as well as the rest of the code from VIX Futures Term Structure R script

```{r}

#Create new data set ####
vix_futures_06_24_2 <- vix_futures_06_24
# clean data 
vix_futures_06_24_2 <- vix_futures_06_24_2 %>% # Remove rows where the value in the "Settle" column is 0
  filter(Settle != 0)


#Calculate Days to Maturity for each contract ####
vix_futures_06_24_2 <- vix_futures_06_24_2 %>%
  mutate(DaysToMat = as.numeric(ExpirationDate - Date))
# order by Days to Maturity 
vix_futures_06_24_2 <- vix_futures_06_24_2[order(vix_futures_06_24_2$DaysToMat), ]







#Create constant Maturity Futures ####

#Define the linear interpolation function
linear_interpolation <- function(df, target_maturity) {
  # Subset the data frame to contracts with maturity greater than the target
  df_above <- subset(df, DaysToMat >= target_maturity)
  # Subset the data frame to contracts with maturity less than the target
  df_below <- subset(df, DaysToMat < target_maturity)
  
  # Find the nearest contracts above and below the target maturity
  contract_above <- df_above[which.min(df_above$DaysToMat), ]
  contract_below <- df_below[which.max(df_below$DaysToMat), ]
  
  # Perform linear interpolation
  tau1 <- contract_below$DaysToMat
  tau2 <- contract_above$DaysToMat
  Ft1 <- contract_below$Settle
  Ft2 <- contract_above$Settle
  
  Ft <- ((tau2 - target_maturity) * Ft1 + (target_maturity - tau1) * Ft2) / (tau2 - tau1)
  
  return(Ft)
}

# Define the target maturities
target_maturities <- c(30, 60, 90, 120, 150, 180)

# Create an empty list to store results
constant_maturity_list <- list()

# Loop over each trading day
for (date in unique(vix_futures_06_24_2$Date)) {
  # Filter contracts traded on the current date
  df_day <- subset(vix_futures_06_24_2, Date == date)
  
  # Sort by days to maturity
  df_day <- df_day[order(df_day$DaysToMat), ]
  
  # Calculate the constant maturity prices for each target maturity
  constant_maturity_prices <- sapply(target_maturities, function(m) {
    linear_interpolation(df_day, m)
  })
  
  # Store the results in the list with the date as the name
  constant_maturity_list[[as.character(date)]] <- constant_maturity_prices
}



# Clean and convert the data
# Calculate the lengths of all elements in constant_maturity_list
elements_lengths <- lengths(constant_maturity_list)

# Find names of the elements with lengths matching the length of target_maturities
correct_length_names <- names(elements_lengths[elements_lengths == length(target_maturities)])

# Subset the original list to include only elements with the correct length
constant_maturity_list_correct <- constant_maturity_list[correct_length_names]

# Convert the corrected list to a data frame
constant_maturity_df <- do.call(rbind, constant_maturity_list_correct)


# Convert the matrix to a data frame
constant_maturity_df <- as.data.frame(constant_maturity_df)
constant_maturity_df <- sapply(constant_maturity_df, function(x) sapply(x, `[`, 1))
constant_maturity_df <- as.data.frame(constant_maturity_df)



# Define maturities corresponding to the V columns
maturities <- target_maturities

# Perform the pivot_longer operation
constant_maturity_long <- pivot_longer(
  constant_maturity_df,
  cols = starts_with("V"),
  names_to = "Maturity",
  values_to = "Price",
  names_prefix = "V",
  names_transform = list(Maturity = function(x) maturities[as.integer(x)])
)


# Convert the 'Price' column to numeric 
constant_maturity_long$Price <- as.numeric(as.character(constant_maturity_long$Price))



# Compute the average
average_term_structure <- constant_maturity_long %>%
  group_by(Maturity) %>%
  summarise(AveragePrice = mean(Price, na.rm = TRUE))


# Add mean spot VIX index level for the same time period 
vix_data_06_24 <- subset(vix_data, Date >= as.Date("2005-05-23") & Date <= as.Date("2024-02-02"))
vix_closing_mean_06_24 <-mean(vix_data_06_24$Close, na.rm = TRUE)

spot_vix_row <- data.frame(Maturity = "Spot VIX", AveragePrice = vix_closing_mean_06_24)

average_term_structure <- rbind(spot_vix_row, average_term_structure)





#Create the Plot for the Term Structure ####
# Convert Maturity to a factor and specify the levels to ensure the correct order
average_term_structure$Maturity <- factor(
  average_term_structure$Maturity,
  levels = c("Spot VIX", "30", "60", "90", "120", "150", "180")
)

# Create the plot with updated styles and larger axis text sizes
term_structure_plot <- ggplot(average_term_structure, aes(x = Maturity, y = AveragePrice)) +
  geom_line(group = 1) +  # Ensure that the line is continuous
  geom_point() +  # Add points for all maturities
  geom_text(
    aes(label = round(AveragePrice, 2)), 
    nudge_y = 0.5, check_overlap = TRUE
  ) +  # Add labels for the points
  scale_x_discrete(  # Use discrete scale for x-axis
    breaks = levels(average_term_structure$Maturity),
    labels = levels(average_term_structure$Maturity)
  ) +
  labs(title = "VIX Futures Term Structure", x = "Days to Maturity", y = "Value") +
  theme_minimal() +  # Use a minimal theme
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", family = "Times New Roman"),  # Center and bold the title in Times New Roman
    axis.title.x = element_text(margin = margin(t = 10), family = "Times New Roman"),  # X-axis title in Times New Roman
    axis.title.y = element_text(margin = margin(r = 10), family = "Times New Roman"),  # Y-axis title in Times New Roman
    axis.text.x = element_text(family = "Times New Roman", size = 12),  # Increase X-axis text size
    axis.text.y = element_text(family = "Times New Roman", size = 12),  # Increase Y-axis text size
    panel.border = element_rect(colour = "black", fill=NA, size=1)  # Add border around the plot
  )

# Print the plot
print(term_structure_plot)










#Create Summary Statistics for VIX Level and constant maturity VIX futures Level####
# Summary Statistics for VIX Futures Levels
# Calculate summary statistics for each maturity level (vfl = vix forward level) 
summary_stats_levels_vfl <- constant_maturity_long %>%
  group_by(Maturity) %>%
  summarise(
    N_Obs = n(),
    Mean = mean(Price, na.rm = TRUE),
    Median = median(Price, na.rm = TRUE),
    StdDev = sd(Price, na.rm = TRUE),
    Skewness = skewness(Price, na.rm = TRUE),
    Kurtosis = kurtosis(Price, na.rm = TRUE),
    Min= min(Price, na.rm = TRUE),
    Max = max(Price, na.rm = TRUE)
  )

print(summary_stats_levels_vfl)

# do the same thing for the spot vix level over the same time period (vsl = vix spot level)
summary_stats_level_vsl <- vix_data_06_24 %>%
  summarise(
    N_Obs = n(),
    Mean = mean(Close, na.rm = TRUE),
    Median = median(Close, na.rm = TRUE),
    StdDev = sd(Close, na.rm = TRUE),
    Skewness = skewness(Close, na.rm = TRUE),
    Kurtosis = kurtosis(Close, na.rm = TRUE),
    Min = min(Close, na.rm = TRUE),
    Max = max(Close, na.rm = TRUE)
  )
summary_stats_level_vsl <- mutate(summary_stats_level_vsl, Maturity = "Spot VIX") #add this column to make it easier to merge the data sets 



# combine data sets
combined_summary_stats_futures <- rbind(summary_stats_levels_vfl, summary_stats_level_vsl)
combined_summary_stats_futures <- combined_summary_stats_futures[c("Maturity", names(combined_summary_stats_futures)[!names(combined_summary_stats_futures) %in% "Maturity"])]


# Get the desired format
# make the data long so that everything except Maturity is in one column
long_data_vix_futures <- pivot_longer(
  combined_summary_stats_futures,
  cols = -Maturity,
  names_to = "Statistic",
  values_to = "Value"
)
# pivot it wider to get the desired format
combined_summary_stats_futures <- pivot_wider(
  long_data_vix_futures,
  names_from = Maturity,
  values_from = Value
)

# Delete first row 
combined_summary_stats_futures <- combined_summary_stats_futures[-1, ]


# Rename the column headers
names(combined_summary_stats_futures)[2:7] <- paste0("VX", c("30", "60", "90", "120", "150", "180"))
# Ensure 'Spot VIX' is the second column
spot_vix_col <- combined_summary_stats_futures[["Spot VIX"]]
combined_summary_stats_futures <- combined_summary_stats_futures[-which(names(combined_summary_stats_futures) == "Spot VIX")]
combined_summary_stats_futures <- cbind(combined_summary_stats_futures[1], Spot_VIX = spot_vix_col, combined_summary_stats_futures[-1])
# Round the values to two decimal places
combined_summary_stats_futures[-1] <- round(combined_summary_stats_futures[-1], 2)
# Delete first column name
names(combined_summary_stats_futures)[1] <- ""
names(combined_summary_stats_futures)[2] <- "Spot VIX"



```

Plot the table for summary stats of VIX and VIX futures Levels

```{r, echo=FALSE, results='asis'}

kable(combined_summary_stats_futures, format = "html", booktabs = TRUE, align = 'c', row.names = FALSE, 
      table.attr = 'class="times-new-roman"') %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
                full_width = FALSE, 
                position = "center",
                font_size = 12) %>%
  column_spec(1, bold = TRUE, color = "black", extra_css = "text-align: left;") %>%
  row_spec(0, extra_css = "border-bottom: 2px solid black;") %>%
  row_spec(7, extra_css = "border-bottom: 3px double black;") %>%
  row_spec(0, extra_css = "border-top: 3px double black;")

```


Code for VIX vs. Futures Returns

```{r, echo=FALSE, results='asis'}

#Create Summary Statistics for VIX Returns and constant maturity VIX futures Returns####

# create new dataset to work on 
constant_maturity_df_dates <- constant_maturity_df
# Convert row names to a column in the dataframe
constant_maturity_df_dates <- rownames_to_column(constant_maturity_df_dates, var = "Date")
# Assuming the "Date" column contains Julian dates and the origin is "1970-01-01"
constant_maturity_df_dates$Date <- as.Date(as.numeric(constant_maturity_df_dates$Date), origin = "1970-01-01")
# Order the dataframe by the "Date" column in ascending order
constant_maturity_df_dates <- constant_maturity_df_dates[order(constant_maturity_df_dates$Date), ]
# Keep only rows with dates on or after "2005-12-22" to get rid of missing values which are created through the creation of the synthetic contracts
constant_maturity_df_dates <- subset(constant_maturity_df_dates, Date >= as.Date("2005-12-22") & Date <= as.Date("2024-01-12"))
# Remove rows with any NA values in constant_maturity_df_dates
constant_maturity_df_dates <- na.omit(constant_maturity_df_dates)
# Convert columns V1-V6 to numeric
constant_maturity_df_dates$V1 <- as.numeric(as.character(constant_maturity_df_dates$V1))
constant_maturity_df_dates$V2 <- as.numeric(as.character(constant_maturity_df_dates$V2))
constant_maturity_df_dates$V3 <- as.numeric(as.character(constant_maturity_df_dates$V3))
constant_maturity_df_dates$V4 <- as.numeric(as.character(constant_maturity_df_dates$V4))
constant_maturity_df_dates$V5 <- as.numeric(as.character(constant_maturity_df_dates$V5))
constant_maturity_df_dates$V6 <- as.numeric(as.character(constant_maturity_df_dates$V6))

# Calculate log returns for each numeric column and add as new columns
constant_maturity_df_dates$VX30 <- c(NA, diff(log(constant_maturity_df_dates$V1)))
constant_maturity_df_dates$VX60 <- c(NA, diff(log(constant_maturity_df_dates$V2)))
constant_maturity_df_dates$VX90 <- c(NA, diff(log(constant_maturity_df_dates$V3)))
constant_maturity_df_dates$VX120 <- c(NA, diff(log(constant_maturity_df_dates$V4)))
constant_maturity_df_dates$VX150 <- c(NA, diff(log(constant_maturity_df_dates$V5)))
constant_maturity_df_dates$VX180 <- c(NA, diff(log(constant_maturity_df_dates$V6)))

# Function to calculate all the statistics
calculate_statistics_vfr <- function(x) {
  return(c(
    Mean = mean(x, na.rm = TRUE),
    Median = median(x, na.rm = TRUE),
    StdDev = sd(x, na.rm = TRUE),
    Skewness = skewness(x, na.rm = TRUE),
    Kurtosis = kurtosis(x, na.rm = TRUE),
    Min = min(x, na.rm = TRUE),
    Max = max(x, na.rm = TRUE)
  ))
}
# Apply the function to each log return column and store the results in a data frame
summary_statistics <- sapply(constant_maturity_df_dates[8:13], calculate_statistics_vfr) # Adjust the column indices accordingly
# Convert the results to a data frame for a better view
summary_stats_returns_vfr <- as.data.frame(summary_statistics) # (vfr = vix forward return)



# Add the VIX data
vix_data_06_24_returns <- vix_data_06_24
vix_data_06_24_returns <- subset(vix_data_06_24_returns, Date >= as.Date("2005-12-22"))

# Calculate summary statistics for the 'Returns' column in 'vix_data_06_24_returns'
calculate_statistics_vsl <- vix_data_06_24_returns %>%
  summarise(
    Mean = mean(Returns, na.rm = TRUE),
    Median = median(Returns, na.rm = TRUE),
    StdDev = sd(Returns, na.rm = TRUE),
    Skewness = skewness(Returns, na.rm = TRUE),
    Kurtosis = kurtosis(Returns, na.rm = TRUE),
    Min = min(Returns, na.rm = TRUE),
    Max = max(Returns, na.rm = TRUE)
  )
# Add a column 'Maturity' with the value 'Spot VIX'
calculate_statistics_vsl <- mutate(calculate_statistics_vsl, Maturity = "Spot VIX")
# Clean the dataset
calculate_statistics_vsl_long <- pivot_longer(
  data = calculate_statistics_vsl, 
  cols = -Maturity, # This selects all columns except for Maturity
  names_to = "Statistic", 
  values_to = "Spot VIX"
)

calculate_statistics_vsl_long <- calculate_statistics_vsl_long %>% 
  select(-Maturity)  # This excludes the Maturity column

calculate_statistics_vsl_long <- calculate_statistics_vsl_long %>%
  column_to_rownames(var = "Statistic")



# Merge the two statistics in one dataset
# If calculate_statistics_vsl_long doesn't have 'Statistic' as a column, create it from rownames
if(!"Statistic" %in% colnames(calculate_statistics_vsl_long)) {
  calculate_statistics_vsl_long <- tibble::rownames_to_column(calculate_statistics_vsl_long, var = "Statistic")
}
# Do the same for summary_stats_returns_vfr
if(!"Statistic" %in% colnames(summary_stats_returns_vfr)) {
  summary_stats_returns_vfr <- tibble::rownames_to_column(summary_stats_returns_vfr, var = "Statistic")
}
# Now merge the datasets
combined_summary_stats_futures_returns <- merge(calculate_statistics_vsl_long, summary_stats_returns_vfr, by = "Statistic", all = TRUE)
# Clean this new dataset 
  # Define the desired order for the "Statistic" column
desired_order <- c("Mean", "Median", "StdDev", "Skewness", "Kurtosis", "Min", "Max")
  # Reorder the dataset based on the desired order
combined_summary_stats_futures_returns <- combined_summary_stats_futures_returns[match(desired_order, combined_summary_stats_futures_returns$Statistic), ]
# Convert specified rows to percentage format 
combined_summary_stats_futures_returns <- combined_summary_stats_futures_returns %>%
  mutate(across(where(is.numeric), ~ ifelse(Statistic %in% c("Mean", "Median", "StdDev", "Min", "Max"),
                                            . * 100, 
                                            .)))
# Round all values (including the percentages) to four decimal places
combined_summary_stats_futures_returns <- combined_summary_stats_futures_returns %>%
  mutate(across(where(is.numeric), ~ round(., 4)))
# For percentage rows, add the "%" sign after rounding
combined_summary_stats_futures_returns <- combined_summary_stats_futures_returns %>%
  mutate(across(where(is.numeric), ~ ifelse(Statistic %in% c("Mean", "Median", "StdDev", "Min", "Max"),
                                            paste0(., "%"), 
                                            .)))

names(combined_summary_stats_futures_returns)[1] <- ""


```


Print the Table for the Returns

```{r, echo=FALSE, results='asis'}

kable(combined_summary_stats_futures_returns, format = "html", booktabs = TRUE, align = 'c', row.names = FALSE, 
      table.attr = 'class="times-new-roman"') %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
                full_width = FALSE, 
                position = "center",
                font_size = 12) %>%
  column_spec(1, bold = TRUE, color = "black", extra_css = "text-align: left;") %>%
  row_spec(0, extra_css = "border-bottom: 2px solid black;") %>%
  row_spec(7, extra_css = "border-bottom: 3px double black;") %>%
  row_spec(0, extra_css = "border-top: 3px double black;")

```




Insert the code for the correlation calculation 
```{r, echo=FALSE, results='asis'}

#Correlation VIX futures and VIX index ####
# Create new dataset based on the returns calculated in the last step 
constant_maturity_df_dates_correlation <- constant_maturity_df_dates
  # Only keep returns 
constant_maturity_df_dates_correlation <- constant_maturity_df_dates_correlation %>%
  select(-V1, -V2, -V3, -V4, -V5, -V6)
  # add returns from spot VIX Index 
    # Perform a left join to match the "Returns" based on the "Date" column
constant_maturity_df_dates_correlation <- left_join(constant_maturity_df_dates_correlation, 
                                                    vix_data_06_24_returns[, c("Date", "Returns")], 
                                                    by = "Date")



# Calculate the actual correlation and create a dataset
  # Initialize a vector to store correlation values
correlations <- c()
  # List of VX column names
vx_columns <- c("VX30", "VX60", "VX90", "VX120", "VX150", "VX180")
  # Calculate the correlation for each VX column with Returns
for (vx_col in vx_columns) {
  corr_value <- cor(constant_maturity_df_dates_correlation[[vx_col]], 
                    constant_maturity_df_dates_correlation$Returns, 
                    use = "complete.obs")
  correlations <- c(correlations, corr_value)
}
  # Create a new data frame with the correlation values
correlation_df <- data.frame(VX_Column = vx_columns, Correlation = correlations)
  # Turn the 'VX_Column' into column names and 'Correlation' into values
transposed_correlation_df <- t(correlation_df[-1])
colnames(transposed_correlation_df) <- correlation_df$VX_Column
  # Convert the matrix to a data frame, if needed
transposed_correlation_df <- as.data.frame(transposed_correlation_df)
  # If you need to make the row names into a proper column
transposed_correlation_df <- tibble::rownames_to_column(transposed_correlation_df, var = "Metric")
  # Rename the value in the 'Metric' column
transposed_correlation_df$Metric <- "Corr. with Spot VIX"
  # Round, multiply, and format all other columns
transposed_correlation_df <- transposed_correlation_df %>%
  mutate(across(-Metric, ~ paste0(format(round(.x * 100, 2), nsmall = 2), "%")))
  # delete the first column name
names(transposed_correlation_df)[1] <- ""
# Print the updated dataset
print(transposed_correlation_df)

```

Create the plot of the corrlation table 

```{r, echo=FALSE, results='asis'}

kable(transposed_correlation_df, format = "html", booktabs = TRUE, align = 'c', row.names = FALSE, 
      table.attr = 'class="times-new-roman"') %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
                full_width = FALSE, 
                position = "center",
                font_size = 12) %>%
  column_spec(1, bold = TRUE, color = "black", extra_css = "text-align: left;") %>%
  row_spec(0, extra_css = "border-bottom: 2px solid black;") %>%
  row_spec(1, extra_css = "border-bottom: 3px double black;") %>%
  row_spec(0, extra_css = "border-top: 3px double black;")

```





Insert the code for the percentage of VIX Futures greater than VIX Level per year 

```{r, echo=FALSE, results='asis'}

#Time Series comparison of the VIX and the VIX Futures ####

# create new dataset to work on as well as clean and format it appropriatly
constant_maturity_df_2 <- constant_maturity_df
# Convert row names to a column in the dataframe
constant_maturity_df_2 <- rownames_to_column(constant_maturity_df_2, var = "Date")

# Assuming the "Date" column contains Julian dates and the origin is "1970-01-01"
constant_maturity_df_2$Date <- as.Date(as.numeric(constant_maturity_df_2$Date), origin = "1970-01-01")


# combine with VIX data 
combined_vix_and_future <- merge(vix_data_06_24, constant_maturity_df_2, by= "Date", all = TRUE)
combined_vix_and_future <- combined_vix_and_future %>%
  rename(
    VX30 = V1,
    VX60 = V2,
    VX90 = V3,
    VX120 = V4,
    VX150 = V5,
    VX180 = V6
  )



# Calculate the percentage where each 'VX' column is greater than the 'Close' column 
# Overall:
percentages_futures_greater <- sapply(combined_vix_and_future[, c("VX30", "VX60", "VX90", "VX120", "VX150", "VX180")], function(vx_column) {
  sum(vx_column > combined_vix_and_future$Close, na.rm = TRUE) / sum(!is.na(vx_column) & !is.na(combined_vix_and_future$Close)) * 100
}, USE.NAMES = TRUE)

# Convert to dataframe 
percentages_futures_greater_df <- as.data.frame(t(percentages_futures_greater))

# Add the 'Year' column with a value that represents the total
percentages_futures_greater_df$Year <- "Total"


# Yearly:  
# Extract the year from the 'Date' column
combined_vix_and_future$Year <- format(combined_vix_and_future$Date, "%Y")

# Initialize a dataframe to hold the percentages for each year (vvf = vix versus futures)
yearly_percentages_vvf <- data.frame()

# Calculate the percentages for each year
for(year in unique(combined_vix_and_future$Year)) {
  # Subset the data for the year
  year_data <- subset(combined_vix_and_future, Year == year)
  
  # Calculate the percentages for the subset
  percentages <- sapply(year_data[, c("VX30", "VX60", "VX90", "VX120", "VX150", "VX180")], function(vx_column) {
    sum(vx_column > year_data$Close, na.rm = TRUE) / sum(!is.na(vx_column) & !is.na(year_data$Close)) * 100
  }, USE.NAMES = TRUE)
  
  # Combine the percentages into a single row data frame with the year as a column
  year_percentages_vvf_df <- data.frame(Year = year, t(percentages))
  
  # Bind the row to the yearly_percentages dataframe
  yearly_percentages_vvf <- rbind(yearly_percentages_vvf, year_percentages_vvf_df)
}

# delete year 2005 due to lack of deta
yearly_percentages_vvf <- yearly_percentages_vvf[-1, ]
# delete year 2024 due to lack of data
yearly_percentages_vvf <- yearly_percentages_vvf[-nrow(yearly_percentages_vvf), ]




# Combine yearly and overall: 
# Bind the total percentages as the last row of yearly_percentages_vvf
yearly_percentages_vvf <- rbind(yearly_percentages_vvf, percentages_futures_greater_df)


# Change col names
names(yearly_percentages_vvf)[names(yearly_percentages_vvf) == "VX30"] <- "VX30>VIX"
names(yearly_percentages_vvf)[names(yearly_percentages_vvf) == "VX60"] <- "VX60>VIX"
names(yearly_percentages_vvf)[names(yearly_percentages_vvf) == "VX90"] <- "VX90>VIX"
names(yearly_percentages_vvf)[names(yearly_percentages_vvf) == "VX120"] <- "VX120>VIX"
names(yearly_percentages_vvf)[names(yearly_percentages_vvf) == "VX150"] <- "VX150>VIX"
names(yearly_percentages_vvf)[names(yearly_percentages_vvf) == "VX180"] <- "VX180>VIX"

# Percentage levels 
yearly_percentages_vvf <- yearly_percentages_vvf %>%
  mutate(across(2:7, ~ paste0(format(round(.x, 2), nsmall = 2), "%")))


```


Create the yearly plot 

```{r, echo=FALSE, results='asis'}

kable(yearly_percentages_vvf, format = "html", booktabs = TRUE, align = 'c', row.names = FALSE, 
      table.attr = 'class="times-new-roman"') %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
                full_width = FALSE, 
                position = "center",
                font_size = 12) %>%
  column_spec(1, bold = TRUE, color = "black", extra_css = "text-align: left;") %>%
  row_spec(0, extra_css = "border-bottom: 2px solid black;") %>%
  row_spec(19, extra_css = "border-bottom: 3px double black;") %>%
  row_spec(18, extra_css = "border-bottom: 1px solid black;") %>%
  row_spec(0, extra_css = "border-top: 3px double black;")

```



Insert the code for Basis and Contango based on VIX regime 
```{r, echo=FALSE, results='asis'}
#The VIX Futures Basis by Volatility Regime ####

# Create new column in Data assigning each Date to a Volatility Regime based on the spot VIX Level 
combined_vix_and_future <- combined_vix_and_future %>%
  mutate(Volatility_Regime = case_when(
    Close <= 20 ~ "VIX<=20",
    Close > 20 & Close <= 30 ~ "20<VIX<=30",
    Close > 30 & Close <= 40 ~ "30<VIX<=40",
    Close > 40 & Close <= 50 ~ "40<VIX<=50",
    Close > 50 ~ "VIX>50"
  ))


# Find the average VIX and VIX Futures Levels for the VIX Regimes
# ensure that all the relevant columns are of numeric type
combined_vix_and_future <- combined_vix_and_future %>%
  mutate(across(c(Close, VX30, VX60, VX90, VX120, VX150, VX180), as.numeric))

# Calculate the averages for each volatility regime
average_values_per_vixregime <- combined_vix_and_future %>%
  group_by(Volatility_Regime) %>%
  summarise(
    Average_Close = mean(Close, na.rm = TRUE),
    Average_VX30 = mean(VX30, na.rm = TRUE),
    Average_VX60 = mean(VX60, na.rm = TRUE),
    Average_VX90 = mean(VX90, na.rm = TRUE),
    Average_VX120 = mean(VX120, na.rm = TRUE),
    Average_VX150 = mean(VX150, na.rm = TRUE),
    Average_VX180 = mean(VX180, na.rm = TRUE)
  )

# Calculate the overall averages across all volatility regimes
overall_averages_vixregime <- combined_vix_and_future %>%
  summarise(
    Average_Close = mean(Close, na.rm = TRUE),
    Average_VX30 = mean(VX30, na.rm = TRUE),
    Average_VX60 = mean(VX60, na.rm = TRUE),
    Average_VX90 = mean(VX90, na.rm = TRUE),
    Average_VX120 = mean(VX120, na.rm = TRUE),
    Average_VX150 = mean(VX150, na.rm = TRUE),
    Average_VX180 = mean(VX180, na.rm = TRUE)
  )

# Add a label for the new row
overall_averages_vixregime <- overall_averages_vixregime %>% mutate(Volatility_Regime = "All Observations")

# Bind the overall averages as a new row to the average_values_per_vixregime dataset
average_values_per_vixregime <- bind_rows(average_values_per_vixregime, overall_averages_vixregime)

# arrange the data 
# Remove the row where Volatility_Regime is NA
average_values_per_vixregime <- filter(average_values_per_vixregime, !is.na(Volatility_Regime))
# manually set "VIX<=20" as the first row and then follow with the rest in ascending order, you can do:
average_values_per_vixregime <- average_values_per_vixregime %>%
  arrange(desc(Volatility_Regime == "VIX<=20")) # This will bring "VIX<=20" to the top



# Calculate the Futures Basis
# Calculate the basis for each VIX futures column
average_values_per_vixregime <- average_values_per_vixregime %>%
  mutate(
    Basis_VX30 = Average_VX30 - Average_Close, 
    Basis_VX60 = Average_VX60 - Average_Close,
    Basis_VX90 = Average_VX90 - Average_Close,
    Basis_VX120 = Average_VX120 - Average_Close,
    Basis_VX150 = Average_VX150 - Average_Close,
    Basis_VX180 = Average_VX180 - Average_Close,
  )

# Calculate the Futures Basis as a % of the VIX 
average_values_per_vixregime <- average_values_per_vixregime %>%
  mutate(
    Basis_VX30_Percent = (Average_VX30 - Average_Close) / Average_Close * 100,
    Basis_VX60_Percent = (Average_VX60 - Average_Close) / Average_Close * 100,
    Basis_VX90_Percent = (Average_VX90 - Average_Close) / Average_Close * 100,
    Basis_VX120_Percent = (Average_VX120 - Average_Close) / Average_Close * 100,
    Basis_VX150_Percent = (Average_VX150 - Average_Close) / Average_Close * 100,
    Basis_VX180_Percent = (Average_VX180 - Average_Close) / Average_Close * 100
  )




# Calculate the % times of Contango
# Calculate the contango percentages for each Volatility Regime
contango_percentages_per_vixregime <- combined_vix_and_future %>%
  group_by(Volatility_Regime) %>%
  summarise(
    Percent_VX30_gt_Close = mean(VX30 > Close, na.rm = TRUE) * 100,
    Percent_VX60_gt_Close = mean(VX60 > Close, na.rm = TRUE) * 100,
    Percent_VX90_gt_Close = mean(VX90 > Close, na.rm = TRUE) * 100,
    Percent_VX120_gt_Close = mean(VX120 > Close, na.rm = TRUE) * 100,
    Percent_VX150_gt_Close = mean(VX150 > Close, na.rm = TRUE) * 100,
    Percent_VX180_gt_Close = mean(VX180 > Close, na.rm = TRUE) * 100
  ) %>%
  ungroup() # Ungrouping so that it can be easily bound with the overall percentages

# Calculate the overall percentages across all observations
contango_overall_percentages <- combined_vix_and_future %>%
  summarise(
    Percent_VX30_gt_Close = mean(VX30 > Close, na.rm = TRUE) * 100,
    Percent_VX60_gt_Close = mean(VX60 > Close, na.rm = TRUE) * 100,
    Percent_VX90_gt_Close = mean(VX90 > Close, na.rm = TRUE) * 100,
    Percent_VX120_gt_Close = mean(VX120 > Close, na.rm = TRUE) * 100,
    Percent_VX150_gt_Close = mean(VX150 > Close, na.rm = TRUE) * 100,
    Percent_VX180_gt_Close = mean(VX180 > Close, na.rm = TRUE) * 100
  ) %>%
  mutate(Volatility_Regime = "All Observations")

# Add the contango percentages to the existing average_values_per_vixregime
average_values_per_vixregime <- average_values_per_vixregime %>%
  left_join(contango_percentages_per_vixregime, by = "Volatility_Regime")
# Remove the last column
contango_overall_percentages <- contango_overall_percentages[, -ncol(contango_overall_percentages)]
# Ensure column names match between the two data frames (they should be the same as per your screenshots)
column_names <- c("Percent_VX30_gt_Close", "Percent_VX60_gt_Close", "Percent_VX90_gt_Close", 
                  "Percent_VX120_gt_Close", "Percent_VX150_gt_Close", "Percent_VX180_gt_Close")
# Replace the NA values in row 6 of 'average_values_per_vix_regime' with the values from 'contango_overall_percentages'
average_values_per_vixregime[6, column_names] <- contango_overall_percentages[1, column_names]









#Create Table for VIX Futures Basis by Volatility Regime ####
# Create a new dataset with selected columns
table_average_values_per_vixregime <- average_values_per_vixregime %>%
  select(Volatility_Regime, Average_Close, Basis_VX30, Basis_VX30_Percent, Percent_VX30_gt_Close, Basis_VX60, Basis_VX60_Percent, Percent_VX60_gt_Close, Basis_VX90, Basis_VX90_Percent, Percent_VX90_gt_Close, Basis_VX120, Basis_VX120_Percent, Percent_VX120_gt_Close)
# rename the columns 
table_average_values_per_vixregime <- table_average_values_per_vixregime %>%
  rename(
    `VIX Level` = Average_Close,
    `Basis 30 Day Futures` = Basis_VX30,
    `Basis 30 Day Futures % of VIX` = Basis_VX30_Percent,
    `% in Contango 30 Day` = Percent_VX30_gt_Close,
    `Basis 60 Day Futures` = Basis_VX60,
    `Basis 60 Day Futures % of VIX` = Basis_VX60_Percent,
    `% in Contango 60 Day` = Percent_VX60_gt_Close,
    `Basis 90 Day Futures` = Basis_VX90,
    `Basis 90 Day Futures % of VIX` = Basis_VX90_Percent,
    `% in Contango 90 Day` = Percent_VX90_gt_Close,
    `Basis 120 Day Futures` = Basis_VX120,
    `Basis 120 Day Futures % of VIX` = Basis_VX120_Percent,
    `% in Contango 120 Day` = Percent_VX120_gt_Close,
  )

# Count the number of observations 
# Count the number of observations for each Volatility_Regime
count_per_vixregime <- combined_vix_and_future %>%
  group_by(Volatility_Regime) %>%
  summarise(Count = n())
# Count the total number of observations
total_count_vix_and_futures <- nrow(combined_vix_and_future)
# Add the total count as a new row in the count_per_vixregime dataset
count_per_vixregime <- count_per_vixregime %>%
  ungroup() %>%
  bind_rows(data.frame(Volatility_Regime = "All Observations", Count = total_count_vix_and_futures))
# Remove the 6th row
count_per_vixregime <- count_per_vixregime[-6, ]
# change the order
desired_top_row_count_per_vixregime <- 4
count_per_vixregime <- count_per_vixregime[c(desired_top_row_count_per_vixregime, 1:(desired_top_row_count_per_vixregime-1), (desired_top_row_count_per_vixregime+1):nrow(count_per_vixregime)), ]


# Put combine the datasets so it has the number of observations in the table_average_values_per_vixregime 
table_average_values_per_vixregime$Count <- count_per_vixregime$Count

# rearrange the order 
column_order_vixregime <- c("Volatility_Regime", "Count", setdiff(names(table_average_values_per_vixregime), c("Volatility_Regime", "Count")))
table_average_values_per_vixregime <- table_average_values_per_vixregime[, column_order_vixregime]

# rename Column 
table_average_values_per_vixregime <- table_average_values_per_vixregime %>%
  rename(`N. Obs.` = Count)


# Make the Table:
table_vixfutures_basis <- kable(table_average_values_per_vixregime, format = "html", align = 'c', caption = "Average Values per VIX Regime") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F) %>%
  column_spec(1, bold = TRUE, border_right = TRUE) %>%
  column_spec(2, width = "50px") # Adjust the width as needed
print(table_vixfutures_basis)





#Create separate Table for Basis only ####
  # Create and rename the columns in the new dataset
table_basis_per_vix_regime <- table_average_values_per_vixregime %>%
  select(
    `Volatility Regime` = Volatility_Regime, 
    `N. Obs.` = `N. Obs.`, 
    `VIX Level` = `VIX Level`, 
    `30 Day Futures Basis` = `Basis 30 Day Futures`, 
    `30 Day Futures Basis as % of VIX` = `Basis 30 Day Futures % of VIX`, 
    `60 Day Futures Basis` = `Basis 60 Day Futures`, 
    `60 Day Futures Basis as % of VIX` = `Basis 60 Day Futures % of VIX`, 
    `90 Day Futures Basis` = `Basis 90 Day Futures`, 
    `90 Day Futures Basis as % of VIX` = `Basis 90 Day Futures % of VIX`, 
    `120 Day Futures Basis` = `Basis 120 Day Futures`, 
    `120 Day Futures Basis as % of VIX` = `Basis 120 Day Futures % of VIX`
  )

  # Round all numerical columns to 2 decimal places, excluding non-numeric columns
table_basis_per_vix_regime <- table_basis_per_vix_regime %>%
  mutate(across(where(is.numeric), ~round(.x, 2)))

#Create separate Table for Contango only ####
# Create and rename the columns in the new dataset
table_contango_per_vix_regime <- table_average_values_per_vixregime %>%
  select(
    `Volatility Regime` = Volatility_Regime, 
    `N. Obs.` = `N. Obs.`, 
    `VIX Level` = `VIX Level`, 
    `30 Day Futures` = `% in Contango 30 Day`, 
    `60 Day Futures` = `% in Contango 60 Day`, 
    `90 Day Futures` = `% in Contango 90 Day`, 
    `120 Day Futures` = `% in Contango 120 Day`, 
  )

# Round all numerical columns to 2 decimal places, excluding non-numeric columns
table_contango_per_vix_regime <- table_contango_per_vix_regime %>%
  mutate(across(where(is.numeric), ~round(.x, 2)))


# Add a % sign to all values in specific columns 
table_contango_per_vix_regime <- table_contango_per_vix_regime %>%
  mutate_at(vars(`30 Day Futures`, `60 Day Futures`, `90 Day Futures`, `120 Day Futures`), 
            ~sprintf("%.2f%%", .))
```



Plot the Basis Table 
```{r, echo=FALSE, results='asis'}
kable(table_basis_per_vix_regime, format = "html", booktabs = TRUE, align = 'c', row.names = FALSE, 
      table.attr = 'class="times-new-roman"') %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
                full_width = TRUE,  # Adjusted to TRUE to make the table wider
                position = "center",
                font_size = 12) %>%
  column_spec(1, bold = TRUE, color = "black", extra_css = "text-align: left;") %>%
  column_spec(1, border_right = "1px solid black") %>%
  column_spec(3, border_right = "1px solid black") %>%
  column_spec(5, border_right = "1px solid black") %>%
  column_spec(7, border_right = "1px solid black") %>%
  column_spec(9, border_right = "1px solid black") %>%
  row_spec(0, extra_css = "border-bottom: 2px solid black;") %>%
  row_spec(6, extra_css = "border-bottom: 3px double black;") %>%
  row_spec(5, extra_css = "border-bottom: 1px solid black;") %>%
  row_spec(0, extra_css = "border-top: 3px double black;")



```


Plot the Contango Table 
```{r, echo=FALSE, results='asis'}


kable(table_contango_per_vix_regime, format = "html", booktabs = TRUE, align = 'c', row.names = FALSE, 
      table.attr = 'class="times-new-roman"') %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
                full_width = FALSE, 
                position = "center",
                font_size = 12) %>%
  column_spec(1, bold = TRUE, color = "black", extra_css = "text-align: left;") %>%
  row_spec(0, extra_css = "border-bottom: 2px solid black;") %>%
  row_spec(6, extra_css = "border-bottom: 3px double black;") %>%
  row_spec(5, extra_css = "border-bottom: 1px solid black;") %>%
  row_spec(0, extra_css = "border-top: 3px double black;")


```